---
title: "ggplot_bio"
output: html_document
---

## R Markdown

```{r}
#Prevent warning and messages from displaying in the pdf of this document
knitr::opts_chunk$set(message = FALSE, warning=FALSE)
```


```{r eval=FALSE}
#eval=FALSE means this won't be run when I knit my final report of this markdown

install.packages("readr")
install.packages("dplyr")
install.packages("ggplot2")
install.packages("cowplot")

```

Load packages
```{r}
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(cowplot)

```

### Read in the data

Data source:

"Coordination of Growth Rate, Cell Cycle, Stress Response, and Metabolic Activity in Yeast"
Brauer *et al*., 2008

http://www.ncbi.nlm.nih.gov/pmc/articles/PMC2174172/

"We studied 36 yeast chemostat cultures growing at six different growth rates under six different nutrient limitations: glucose, sulfate, phosphate, ammonium, leucine (in a nonreverting leu2 mutant), and uracil (in a nonreverting ura3 mutant)"

```{r}
#This sets a display option for tables that lets you see all the columns at once
options(dplyr.width = Inf)

raw_data <- read_delim("ggplot_dataset_brauer2008.txt", delim = "\t" )

```

### Changing to a tidy format

Instead of one gene per row, we want one observation of that gene per row

The same information about the sky in wide form and in tidy form: A Metaphor

*Wide form data*

 - The sky has colors and august it was blue but this other time at sunset it was red but in the winter it's grey

*Tidy form data*

 - The sky in August was blue
 - The sky in December was grey 
 - The sky at Sunset was red

ie. 

 - gene experiment_condition1 value1
 - gene experiment_condition2 value2


```{r}
# Here I'm changing the data to tidy
# Here's what the incoming dataframe looks like

head(raw_data)

# I'm using the gather function from dplyr to do this
# The values starts in column #5 and go to the end.
# I want all of these values to go in a column called "expression"
# I want the old column headers to go in a column called "condition"
tidy_data <- raw_data %>% gather(key = condition, value = expression, 5:ncol(raw_data))

#Equivalent expression
#tidy_data <- gather(data = df_data, key = condition, value = expression, 5:ncol(raw_data))

head(tidy_data)

# As a rule, each cell should only every hold one value/attribute

# Some columns contain multiple different values ex. SFB2::ER to Golgi transport::molecular function unknown::YNL049C::1082129
# We will split this up into columns

#I'm using the separate function to take the column NAME, and split in on the separator ":", giving each columns new names
yeast_tmp <- tidy_data %>% separate(col = NAME, into = c("gene", "GO_bio", "GO_mol", "alt_name", "position"), sep="::" )

head(yeast_tmp)

#The experimental condition column has nutrient and rate in the same column. 
#Split these up as well

yeast <- yeast_tmp %>% separate(condition, c("nutrient", "rate"), sep=1)
head(yeast)

#I want the rate column to be numeric, not a string
yeast$rate <- as.numeric(yeast$rate)
head(yeast)

#Here, I'm replacing the single letter nutreint code with the full word
#The grammar of this sentence is:
#In every position where yeast$nutrient equals "G", replace the position's entry with "Glucose"
yeast$nutrient[yeast$nutrient == "G"] <- "Glucose"
yeast$nutrient[yeast$nutrient == "L"] <- "Leucine"
yeast$nutrient[yeast$nutrient == "N"] <- "Ammonium"
yeast$nutrient[yeast$nutrient == "P"] <- "Phosphate"
yeast$nutrient[yeast$nutrient == "S"] <- "Sulfate"
yeast$nutrient[yeast$nutrient == "U"] <- "Uracil"


yeast <- yeast %>% filter(gene != "")

```

### Demo of some ggplot geoms

A geom is a type of plot, like scatter, violin, bar, density, etc. 

List of the current ggplot geoms:

http://docs.ggplot2.org/current/


```{r}

#I'm taking a random sample of rows from this dataset to demo plots
#If plots take a long time to render, subsampling can initially helpful

#Sample 20,000 rows from the yeast dataframe
miniyeast = sample_n(yeast, 10000)


# Each ggplot is like a recipe where you add on things you want
# Start by initializing the ggplot with what data you want to use in the plot
# the 'aes' function is where you assign what columns to use for each plot feature
ggplot(data=miniyeast, aes(x=rate, y=expression, color=nutrient)) +
   geom_point() 

# Jitter spreads out points that have the same x coordinate
ggplot(data=miniyeast, aes(x=rate, y=expression, color=nutrient)) +
   geom_jitter() 

# Here, I don't want all the points for each condition on to of each other
# I'm adding on 'facet_wrap' by the variable 'nutrient' to get one
# panel for each nutrient

ggplot(data=miniyeast, aes(x=rate, y=expression, color=nutrient)) + 
   geom_jitter() +  
   facet_wrap(~nutrient)  

# 
ggplot(data=miniyeast, aes(x=rate, y=expression, color=nutrient)) +
   geom_boxplot() +
   facet_wrap(~nutrient)

ggplot(data=miniyeast, aes(x=rate, y=expression, group=rate, color=nutrient)) +
   geom_boxplot() +
   facet_wrap(~nutrient)

ggplot(data=miniyeast, aes(x=rate, y=expression, group=rate, color=nutrient)) +
   geom_violin() +
   facet_wrap(~nutrient)

ggplot(data=miniyeast, aes(expression, color=nutrient)) +
   geom_density() +
   facet_wrap(~nutrient)


#Take a look at all the the data together
#Takes longer to plot, but isn't unreasonable
ggplot(data=yeast, aes(x=rate, y= expression, group=gene, color=nutrient)) +
   geom_line() +
   theme(legend.position="none") +  #We don't want a giant legend with each gene name
   facet_wrap(~nutrient)


```

#### Doing an analysis and visualizing the results

 - I saw that there are tracks which have positive and negative slopes in all the nutrients
 
 - I decided to look for genes which are up in some conditions and down in others using the dplyr functions from last week
 
 - I take the pearson correlations of rate vs. expression for each gene:nutrient pair 
 
 - I then take the genes which have high variance between their pearson correlations for the different nutrients
```{r}

yeast %>%
   group_by(gene, nutrient) %>% 
   dplyr::summarise(Pearson = cor(rate,expression)) %>% #make a new column called "Pearson" with the correlation btw rate and expression
   group_by(gene) %>%                                   #Regroup by gene
   dplyr::summarize(variance = var(Pearson)) %>%        #Take the variance of each genes correlations in the different nutrients
   filter(variance > 0.75) -> variable_gene_list        #Save genes with high variance

print(variable_gene_list)

# Pull the data from the full dataset for the set of genes with high variance
variable_data <- yeast %>% filter(gene %in% variable_gene_list$gene)

ggplot(data=variable_data  , aes(x=rate, y=expression, group=nutrient, color=nutrient)) +
    geom_line() + 
    facet_wrap(~gene)

# I noticed that there is a set of genes which are down in glucose/leucine and up in the other conditions
# Let's pull just those out to make a figure
selected_genes <- yeast %>% filter(gene %in%  c("ARO3", "DIC1", "ECM40", "TMT1", "TRP3", "VHT1"))

ggplot(data=selected_genes , aes(x=rate, y=expression, group=nutrient, color=nutrient)) +
    geom_line() + 
    facet_wrap(~gene)

```


###Make it look nice

Getting the data accurately displayed is just the first step of plotting

The default plot always needs to be cleaned up for publication/presentation

```{r}

#custom palette
palette <- c("#0072B2","#E69F00","#009E24","#FF0000", "#979797","#5530AA")

#loading cowplot increases text size and removes the default grey background
#It is also really useful for more advanced plotting later on

#Here, I'm adding and silencing features I want fir the final plot
#I find plot features by googling something to the tune of:
#      change line thickness geom_line ggplot
#      ggplot use custom color palette

pairs_plt <- ggplot(data=selected_genes , aes(x=rate, y=expression, group=nutrient, color=nutrient)) +
     geom_line(size=1.5, alpha=0.9) +                                 # Thicker lines, and slightly transparent
     facet_wrap(~gene) +                                              # Each gene gets a plot
     theme(strip.background=element_blank()) +                        # I don't want the grey box around the gene IDs
     theme(axis.text.x=element_text(angle=45, vjust =1, hjust = 1)) + # I want the x axis labels at at 45 degree angle
     ylab("Expression") +                                             # New y label
     xlab("Growth rate") +                                            # New x label
     scale_color_manual(values=palette)                               # Use custom colors

pairs_plt

```

### An aside about colors and plotting

This plot uses a palette of colors I prefer over the default

Default ggplot pastels are *not* colorblind accessible

 - Go for bold, saturated colors, avoid pastels and anything dull/washed-out
 - Use thick lines, particularly in powerpoint presentations
 - Try to use redundant encoding for conditions (like shapes, dashed vs. solid etc.) 
 - Avoid non-intuitive problematic color pairings like:
 
         teal/white
         
         lightpink/grey
         
         red/black
         
         darkblue/purple
         
         dull green/dull orange
         
 - Do not use thin red and black lines or small red/black dots to represent different data conditions
 - Make heatmaps blue/yellow, NOT red/green

### Another plot
#### And a dire warning about using bar charts

```{r}
# Picking genes that have are really high pearson correlations between rate and expression in at least one conditions

yeast %>% 
  group_by(gene, nutrient) %>% 
  dplyr::summarise(Pearson = cor(as.numeric(rate),expression)) %>% 
  filter (Pearson < -0.99) %>% 
  tail(50) -> negcor_genes

print(negcor_genes)
#Selecting the genes I found with high position correlation from the full dataset
negcor <- yeast %>% 
     filter(gene %in% negcor_genes$gene) 

negplt <- ggplot(data=negcor , aes(x=rate, y=expression)) +
     geom_violin() +
     geom_jitter(alpha=0.3)+
     theme(axis.text.x=element_text(angle=45, vjust =1, hjust = 1)) +
     scale_color_manual(values=palette) 
negplt

# Aes can be assigned to individual plots
negplt <- ggplot(data=negcor , aes(x=rate, y=expression)) +
     geom_violin(aes(group=rate)) +
     geom_jitter(alpha=0.3)+
     theme(axis.text.x=element_text(angle=45, vjust =1, hjust = 1)) +
     scale_color_manual(values=palette) 

negplt

#Boxplot version
negplt_box <- ggplot(data=negcor , aes(x=rate, y=expression, group=rate)) +
     geom_boxplot(alpha=0.3)+
     theme(axis.text.x=element_text(angle=45, vjust =1, hjust = 1)) +
     scale_color_manual(values=palette) 

negplt_box 

```

### Plotting the exact same data as above as a bar plot
There are definitely times when bar plots are appropriate

And then there's most other times

Bar plots hide the underlying distribution of observations in data

 - Many papers represent data as bar plot + standard error of the mean (SEM)
 - Standard error is almost never the appropriate statistic, but people use it because it is small and makes plots look more impressive
 - Standard error of the means tells almost nothing about the underlying distribution of the data
 - Choosing a plot format because it hides something troubling in the data => unethical
 - If a violin, jitter, or box plot can be used, it's almost 100% the preferable way to present the data
 - Once you start to look for bar plots +/- SEM in papers, you find them everywhere

```{r}

# Don't make plots like this
# Don't use standard error of the mean to hide variation
negplt_mean<- ggplot(data=negcor , aes(x=rate, y=expression)) +
     geom_bar(stat="summary", fun.y="mean") +
     stat_summary(fun.data = mean_se, geom = "errorbar") +
     theme(axis.text.x=element_text(angle=45, vjust =1, hjust = 1)) +
     scale_color_manual(values=palette) +
     annotate ("text", x = 0.2, y=0.8, label="NO", color="red", size=15)

negplt_mean

# This is not an accurate representation of the data compared to the previous two charts

```

### Assembling and saving a figure
```{r}

final_figure <- plot_grid(pairs_plt, negplt, labels = c("A", "B"), ncol = 2, rel_widths = c(1,0.6))

final_figure

ggsave(plot = final_figure, file = "Figure1.pdf", device= "pdf", width=7, height=5, units="in")

```



